---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# elucidate

<!-- badges: start -->
<a id="devex-badge" rel="Exploration" href="https://github.com/BCDevExchange/assets/blob/master/README.md"><img alt="Being designed and built, but in the lab. May change, disappear, or be buggy." style="border-width:0" src="https://assets.bcdevexchange.org/images/badges/exploration.svg" title="Being designed and built, but in the lab. May change, disappear, or be buggy." /></a>
<!-- badges: end -->

## Project Status

This package is currently maintained by [Craig Hutton](https://craig.rbind.io/), a Data Science Fellow working with the Data Science and Analytics Branch of the [British Columbia Ministry of Citizens’ Services](https://www2.gov.bc.ca/gov/content/governments/organizational-structure/ministries-organizations/ministries/citizens-services) and the Research Branch of the [British Columbia Ministry of Social Development & Poverty Reduction](https://www2.gov.bc.ca/gov/content/governments/organizational-structure/ministries-organizations/ministries/social-development-poverty-reduction).

## Why `elucidate`?

`Elucidate` provides a collection of convenience functions to make data analysis in R easier and more accessible for researchers:

* Functions that help summarize data of multiple types (numeric, character strings, logicals, etc.) using descriptive statistics (`describe*`) split by any number of grouping variables with comparable performance to existing alternatives and enhanced customization via arguments.

* Functions that make it easier to access the power of the [boot](https://cran.r-project.org/web/packages/boot/index.html) package to obtain confidence intervals for a variety of summary statistics (`*_ci`), with options for within-group calculations and parallelized computation to maximize performance.

* Functions that help you visualize your data using the popular [ggplot2](https://ggplot2.tidyverse.org/) & [plotly](https://plot.ly/r/) packages more easily via a relatively simple API with sensible and well documented arguments rather than layers (`plot_*`). 

* Miscellaneous functions that fill in a few of the (rare) gaps in `R`'s statistical computing & data science toolkit, e.g. functions which are easy for experienced users to write on the fly but are challenging for `R` novices or researchers without programming experience to implement: the standard error of the mean (`se`), an operator that returns the values that do not match a vector (`%ni%`, i.e. the opposite of `%in%`), the mode of a numeric variable (`mode_of_y`), etc.

Inspired by tidyverse naming conventions, the core functions of `elucidate` are organized into sets that begin with a common root (e.g. `describe*`, `plot_*`), since this enables the user to see them all as suggestions as you are coding in R studio.

Drawing from similar inspiration, many elucidate functions are also designed to accept a data object as the 1st argument and to return a data or plotting object (e.g. ggplot2 or plotly) so they are compatible with the pipe operator from the [magrittr](https://magrittr.tidyverse.org/reference/pipe.html) package for easy integration into data processing pipelines.

## Installation

You can install the development version of elucidate from the [BC SDPR Research Branch Azure repository](https://dev.azure.com/BC-SDPR-Research/R-elucidate) with:


``` r
# use devtools to install from a git repository
library(devtools)

install_git("git@ssh.dev.azure.com:v3/BC-SDPR-Research/R-elucidate/R-elucidate", 
dependencies = TRUE)

# if you need to specify a non-standard SSH keyfile
cred <- git2r::cred_ssh_key('c:\\.ssh\\id_key.pub', 'c:\\.ssh\\id_key')

install_git(
  "git@ssh.dev.azure.com:v3/BC-SDPR-Research/R-elucidate/R-elucidate",
  credentials = cred,
  dependencies = TRUE
)

# if you want to install a specific branch or tag (e.g. a tagged version 1.0)
install_git(
"git@ssh.dev.azure.com:v3/BC-SDPR-Research/R-elucidate/R-elucidate",
branch = "v1.0",
dependencies = TRUE
)
``` 

The authors of `elucidate` acknowledge and express their gratitude to the authors of the [tidyverse packages](https://www.tidyverse.org/), [data.table](https://github.com/Rdatatable/data.table), and the functions of other dependency packages which were used to build `elucidate`, since without their effort and ingenuity `elucidate` would mostly have remained a collection of ideas instead of functions. 

## Usage

Use `describe()` to describe a single variable in a data frame of vector of values:

```{r describe, cache = TRUE}
library(elucidate)

#using a vector as input
describe(data = rnorm(1:1000, 100, 5))

#using a data frame and specifying a variable in that data frame

describe(data = pdata, y = y1) #no grouping variables, numeric input class 

describe(pdata, g) #factor input class 

describe(pdata, even) #logical input class 

pdata %>% describe(high_low) #character input class, data piped in

describe(pdata, d) #date input class 

#obtain within-group summaries by adding grouping variables

describe(pdata, y1, high_low) #one grouping variable

describe(pdata, y1, high_low, g) #two grouping variables

description <- describe(pdata, y1, high_low, g, even) #three grouping variables

```

To describe all variables in a data frame, use `describe_all()`:

```{r describe_all}

#if more than one class of variable is summarized you'll get a list
#if only one class is present in the data, or only one is requested, you'll get a dataframe

describe_all(pdata) #all summary types in a list, not split by any grouping variables

#only want summaries of certain variable classes? No problem, use the class argument: 

#numeric summary only, split by a grouping variable
pdata %>% 
  describe_all(high_low, 
               output = "dt", #you can ask for a tibble or data.table as output options
               class = "n") 

#numeric and logical summaries only, split by a grouping variable
pdata %>%
 describe_all(high_low, output = "dt", class = c("n", "l"))

```


If you want the counts for unique values of a variable, regardless of class, use `counts()` or `counts_all()`:

```{r counts}

counts(pdata$high_low) #output stucture is value_count, sorted in descending order by default

#use the order argument to sort in ascending order instead
counts(pdata$high_low, order = "a") 

counts_all(pdata[, c("high_low", "even", "g")])

```


Calculate confidence intervals for a sample mean, median or other summary statistic using `describe_ci()` & `describe_ci_all()`:

```{r ci}

#confidence intervals for the mean are calculated (by default) using a theoretical normal distribution...
pdata[1:100, ] %>% describe_ci(y1, stat = mean)

pdata[1:100, ] %>% describe_ci(y1, g, stat = mean) #obtain CIs and means split by a grouping variable

#confidence intervals for other statistics are obtained using bootstrapping
pdata[1:100, ] %>% 
  describe_ci(y1, stat = median) #bootstrapped median

pdata[1:100, ] %>% 
  describe_ci(y1, stat = sd) #bootstrapped standard deviation

#describe_ci_all will return CIs for all numeric variables in a data frame

describe_ci_all(pdata[1:1000, ], stat = median) #bootstrapped CIs for the median

describe_ci_all(pdata, stat = mean) #the default

describe_ci_all(pdata, high_low, stat = mean) #split by a grouping variable

```



Use the `plot_*` set of functions to easily obtain ggplot2 or plotly visualizations:

```{r plot}
data(mtcars)

#basic density plot
plot_density(mtcars, x = mpg)

#histogram
plot_histogram(mtcars, x = mpg, fill_var = cyl) #mapping of variables to aesthetics is accomplished using *_var arguments

#customized density plot
mtcars %>% plot_density(x = mpg,
                        colour_var = cyl, #assign a variable to colour
                        colour_var_order = c("6", "8", "4"), #reorder the levels of the colour variable
                        colour_var_labs = c("six" = "6", "eight" = "8"), #recode the colour variable labels
                        colour_var_values = c("blue3", "red3", "green3"), #change the colours from the ggplot2 defaults
                        colour_var_title = "# cylinders")


#boxplot showing additional customization
plot_box(data = mtcars, #data source
         x = cyl, #variable on the x-axis
         y = hp, #variable to go on the y-axis
         xlab = "# of cylinders", #custom x-axis label
         ylab = "horsepower", #custom y-axis label
         fill_var = am, #assign variable am to fill
         fill_var_order = c("0", "1"), 
         fill_var_title = "transmission", #relabel the fill variable title in the legend
         fill_var_labs = c("manual" = "0", "automatic" = "1"), #recode the fill variable labels
         fill_var_values = c("blue", "red"), #change the colours from the ggplot2 defaults
         theme = "bw") #specify the theme

#violin plot with quantiles added
mtcars %>% plot_violin(y = mpg, x = cyl, fill = "blue", draw_quantiles = c(0.25, 0.5, 0.75))

#scatterplot with a regression lines
mtcars %>% plot_scatter(y = mpg, x = hp, colour_var = cyl, 
                        regression_line = T, 
                        regression_method = "lm") #default is to fit lines using a generalized additive model

#polynomial regression line can be added by specifying a model formula
mtcars %>%
  plot_scatter(y = mpg, x = hp,
               regression_line = T, regression_method = "lm", regression_se = T,
               regression_formula = y ~ poly(x, 2))


#plot a statistic with error bars
library(gapminder) #import the gapminder dataset from the gapminder package

gapminder %>%
   plot_stat_error(y = lifeExp, x = year, 
                   stat = "mean", error = "se", #mean +/- standard error, default error metric is a 95% CI
                   colour_var = continent, 
                   geom = "point", #either "bar" or "point"
                   p_size = 3, #adjust the size of the points
                   add_lines = T, #connect the points with lines. This is useful for repeated-measures data.
                   alpha = 0.6) #adjusts the transparency

```

The `%ni%` operator ("ni" for "not in") can help you subset your data like the `%in%` operator but returns an indicator for non-matches instead of matches. This is particularly useful if you want to specify exclusion criteria instead of inclusion criteria.

```{r, eval = FALSE}
#subset data to extract rows with matching values using "%in%"
subset(pdata, g %in% c("a", "e"))

#subset data to extract rows with non-matching values using "%ni%"
subset(pdata, g %ni% c("a", "e"))

#equivalent to subset function for tidyverse users
dplyr::filter(pdata, g %ni% c("a", "e"))

#output omitted for space considerations
```

## Development Notes

`Elucidate` is still in the formative stage of development and considerable changes may occur to it over the next several months leading up to public release.

## Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please contact the package maintainer by [e-mail](mailto:Craig.Hutton@gov.bc.ca?subject=elucidate). After `elucidate` has been released publicly, a more formal method of reporting issues and requests will be implemented.

## License

Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the “License”); you may not use this file except in compliance with the License. You may obtain a copy of the License at

[http://www.apache.org/licenses/LICENSE-2.0](http://www.apache.org/licenses/LICENSE-2.0)

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an “AS IS” BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied. See the License for the specific language governing permissions and limitations under the License.

